 Solving ODEs in R

### 2014-11-20 documentation/solving_ODEs.Rmd
solving ODEs: add simulation of gradient, add references

---
title: " Solving ODEs in R"
author: "Florian Schneider"
date: "Thursday, November 20, 2014"
output:
  html_document:
    keep_md: yes
  pdf_document:
    fig_height: 3.5
    fig_width: 4
fontsize: 12pt
geometry: margin=1.4in
csl: ../manuscript/amnat.csl
bibliography: ../manuscript/cas02.bib
---

@@

In R, the package 'deSolve' provides solvers for differential equations. 
```{r}
library("deSolve")
```

@@

We need to define the constants, as well as the initial vegetation structure which is unambigously given by the values $\rho_{+-}$, $\rho_{++}$, $\rho_{--}$, $\rho_{+}$ and $\rho_{-}$. We define the starting values as a static function of $\rho_{+}$ and store all of them in a single vector called `rho` which is expected by the functions defined above.

```{r}
# model parameters
model_parms <- list(
  delta = 0.1, # proportion of seeds dispersed globally
  b = 0.75, # environmental quality
  m = 0.15, # plant mortality
  f = 0.9, # effect of abiotic facilitation on regeneration
  r = 0.01, # intrinsic regeneration 
  d = 0.1, # rate of degradation
  c = 0.2 # global competition among plants 
  )

# initial vegetation cover
rho_ini <- 0.2

@@

```{r}
# running the ode-solver
runmodel <- ode(y = rho, func = odesys, times = 1.1^(1:60), parms = model_parms)

# transfer into ouput and calculate missing rho values
out <- as.data.frame(runmodel)
out$rho_0 <- 1 - out$rho_p - out$rho_m
out$rho_p0 <- with(out, rho_p-rho_pm-rho_pp)
out$rho_m0 <- with(out, rho_m-rho_pm-rho_mm)
out$rho_00 <- with(out, rho_0-rho_m0-rho_p0)

# plot 
plot(rho_p ~ time, data = out, type = "l", lwd = 2, ylab = "rho_plus", main = "pair-approximation", ylim = c(0,1))
lines(rho_m ~ time, data = out, lty = 3)
lines(rho_0 ~ time, data = out, col = "grey50")
```

## simulating over given $\rho_+$
For the project, we need a prediction of growth for a given value of $\rho_+$. To achieve this, I am simulating a gradient of replicates with different environmental conditions.

```{r, echo = FALSE}
library(foreach)
library(doSNOW)

@@

@@

global <- list(  
  m = 0.1, #intrinsic mortality
  d = 0.1,    # degradation
  c = 0.2, 		# beta*g  
  delta = 0.1,  	# seeds dispersed; (1-del) seeds on nearest neighbourhood
  r = 0.01, 	# regeneration rate
  f = 0.9		# local fascilitation
)

parameters <- list(
  b = seq(0,1,.01) #seq(0,1,.002)+0.001
)

iterations <- expand.grid(parameters)
iterations <- cbind(ID = 1:dim(iterations)[1],iterations, global)
str(iterations)

# initial vegetation cover
rho_ini <- 0.2

@@

  # running the ode-solver
  runmodel <- ode(y = rho, func = odesys, times = c(1,60), parms = model_parms)

@@

  return(out[dim(out)[1],])
  } -> output

plot(output$rho_p ~ iterations$b, type = "l", ylim = c(0,1))

stopCluster(cl)

>>>>

---
title: ' Solving ODEs in R'
author: "Florian Schneider"
date: "Thursday, November 20, 2014"
output:
  html_document:
    keep_md: yes
  pdf_document:
    fig_height: 5
    fig_width: 7
fontsize: 10pt
geometry: margin=1.2in
csl: ../manuscript/amnat.csl
bibliography: ../manuscript/cas02.bib
---

@@

In R, the package 'deSolve' provides solvers for differential equations [@soetaert10].

```{r}
library("deSolve")
```

@@

**Important:** *Equations of ODE and function for ODE solver are saved in file `code/ode_kefi.r` for further use in simulations and applications.*

We need to define the constants, which I take from the example of @kefi07, Fig. 3 to evaluate this implementation (looks very similar, differences might originate from the starting conditions, which are not given precisely in the paper).

```{r}
# model parameters
model_parms <- list(
  delta = 0, # proportion of seeds dispersed globally
  b = 0.4, # environmental quality
  m = 0.1, # plant mortality
  f = 0.8, # effect of abiotic facilitation on regeneration
  r = 0.0, # intrinsic regeneration 
  d = 0.2, # rate of degradation
  c = 0.3 # global competition among plants 
  )
```

And the initial vegetation structure which is unambigously given by the values $\rho_{+-}$, $\rho_{++}$, $\rho_{--}$, $\rho_{+}$ and $\rho_{-}$. We define the starting values as a static function of $\rho_{+}$ and store all of them in a single vector called `rho` which is expected by the functions defined above.

```{r}
# initial vegetation cover
rho_ini <- 0.1

@@

Note that I am determining the timesteps exponentially, to have a decline in resolution as the dynamics steady out.

```{r}
# running the ode-solver
runmodel <- ode(y = rho, func = odesys, times = 1.1^(1:53), parms = model_parms)

# transfer into ouput and calculate missing rho values
out <- as.data.frame(runmodel)
out$rho_0 <- 1 - out$rho_p - out$rho_m
out$rho_p0 <- with(out, rho_p-rho_pm-rho_pp)
out$rho_m0 <- with(out, rho_m-rho_pm-rho_mm)
out$rho_00 <- with(out, rho_0-rho_m0-rho_p0)
```
```{r, "fig1"}
# plot
plot(rho_p ~ time, data = out, type = "l", lwd = 2, ylab = "vegetation cover", ylim = c(0,1))
lines(rho_m ~ time, data = out, lty = 3)
lines(rho_0 ~ time, data = out, col = "grey50")
legend(60,1, c("rho_+", "rho_-", "rho_0"), col = c("black", "black", "grey50"), lty = c(1,3,1), lwd = c(2,1,1), bty = "n", horiz = TRUE)
```

## simulating gradients

For the project, we need a prediction of growth for a given value of $\rho_+$. To achieve this, I am simulating a gradient of replicates with different environmental conditions.

First, we need to generate a parallel backend for the use of the `foreach()` function.

```{r, }
library(foreach)
library(doSNOW)

@@

Then, build a database of parameters, using a global dataset (valid for all iterations) plus a gradient over $b$, stored in `parameters` (I use `expand.grid()` to blow up the parameter space if varying multiple parameters at a time).

@@

global <- list(  
  m = 0.2, #intrinsic mortality
  d = 0.1,    # degradation
  c = 0.2, 		# beta*g  
  delta = 0.9,  	# seeds dispersed; (1-del) seeds on nearest neighbourhood
  r = 0.01, 	# regeneration rate
  f = 0.9		# local fascilitation
)

parameters <- list(
  b = seq(0,1,.0025) #seq(0,1,.002)+0.001
)

iterations <- expand.grid(parameters)
iterations <- cbind(ID = 1:dim(iterations)[1],iterations, global)

In a first simulation run, the initial vegetation cover should be high.

```{r}
# initial vegetation cover
rho_ini <- 0.8

Then, call `foreach()` to run all iterations in parallel.

```{r}
foreach(iteration = iterations$ID, .combine = rbind, .packages = c("deSolve")) %dopar% { 
 model_parms <- as.list(iterations[iteration,])

  # running the ode-solver
  runmodel <- ode(y = rho, func = odesys, times = c(1,10000), parms = model_parms)

  return(out[dim(out)[1],])
  } -> output_high

In a second run, we initialise the same set of parameters, but starting from low cover:

```{r}
# initial vegetation cover
rho_ini <- 0.01

@@

  # running the ode-solver
  runmodel <- ode(y = rho, func = odesys, times = c(1,10000), parms = model_parms)

@@

  return(out[dim(out)[1],])
  } -> output_low
```

```{r, "fig2"}
plot(output_high$rho_p ~ iterations$b, type = "p", pch = 20, cex =0.75, ylim = c(0,1), xlab = "environmental quality, b", ylab = "vegetation cover, rho_+")
points(output_low$rho_p ~ iterations$b, pch = 20, cex =0.75)
```

Don't forget to shut down the parallel backend.
```{r}
stopCluster(cl)
```
